---
title: "useR to programmeR"
subtitle: "Functions 2"
author: "Emma Rand and Ian Lyttle"
format: 
  revealjs:
    theme: [simple, styles.scss]
    slide-number: true
    chalkboard: true
    code-link: true
    code-line-numbers: false
bibliography: references.bib
---

## Learning objectives

In this session, we will discuss:

:::{.incremental}
  - embracing `{{}}` for `<data-masking>` functions
  - "standard" style and design of functions
  - the joys of side-effects
:::

. . .

<hr>

For coding, we will use `r-programming-exercises`:

 - `R/functions-02-01-embrace.R`, etc.

## Plot functions: Motivation 

```{r}
library("tidyverse")
```

Sometimes, you want to generalize a certain type of plot - let's say a histogram:

```r
diamonds |> 
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

diamonds |> 
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.05)
```

"I want choose only the data, variable, and bin-width"

## Function

`aes()` is a data-masking function; you can embrace 🤗

```{r}
histogram <- function(df, var, binwidth = NULL) {
  df |> 
    ggplot(aes(x = {{ var }})) + 
    geom_histogram(binwidth = binwidth)
}

diamonds |> histogram(carat, 0.1) 
```

## Your turn

Complete this function yourself:

```r
histogram <- function(df, var, binwidth = NULL) {
  df |>
    ggplot(aes()) +
    geom_histogram(binwidth = binwidth)
}
```

- Try with other `df` and `var`, e.g. `starwars`, `mtcars`.

- Using *this* function *as is*, how can you:

  - add `theme_minimal()`?

  - fill the bars with `"steelblue"`?

## Your turn (solution)

Adding a theme: `histogram()` returns a `ggplot` object, so you can add a theme in the "usual" way:

```{r}
histogram(starwars, height) + theme_minimal()
```

## Your turn (solution)

*As is*, there is no easy way to specify `"steelblue"`.

. . .

However, you can build an escape hatch.

`...` are called "dot-dot-dot" or "dots".

. . .

```r
# `...` passed to `geom_histogram()`
histogram <- function(df, var, ..., binwidth = NULL) {
  df |>
    ggplot(aes(x = {{var}})) +
    geom_histogram(binwidth = binwidth, ...)
}
```

Passes unspecified arguments from your function to another (tell your users where).

. . .

[Tidyverse Design Guide](https://design.tidyverse.org/dots-after-required.html) has more details.

## Your turn (continued)

Incorporate dot-dot-dot into `histogram()`:

```r
histogram <- function(df, var, ..., binwidth = NULL) {
  df |>
    ggplot(aes(x = {{var}})) +
    geom_histogram(binwidth = binwidth, ...)
}
```

Try, e.g.:

```r
histogram(starwars, height, binwidth = 5, fill = "steelblue")
```

## Tradeoffs


You write a function to make some things easier.

. . .

The cost is that some things become more difficult.

. . .

This is unavoidable, the best you can do is be deliberate about what you make easier and more difficult.


## Labelling 

How to build a string, using variable-names and values?

. . .
  
`rlang::englue()` was built for this purpose:

  - embrace 🤗 variable-names: `{{}}`
  - glue values: `{}`
  
. . .

```{r}
temp <- function(varname, value) {
  rlang::englue("You chose varname: {{ varname }} and value: {value}")
}

temp(val, 0.4)
```

## Your turn

Adapt `histogram()` to include a title that describes:

  - which `var` is binned, and the `binwidth`

```r
histogram <- function(df, var, ..., binwidth = NULL) {
  df |> 
    ggplot(aes(x = {{ var }})) + 
    geom_histogram(binwidth = binwidth, ...) + 
    labs(
      title = rlang::englue("")
    )
}
```

Try:

```r
histogram(starwars, height, binwidth = 5)
histogram(starwars, height) # "extra credit"
```

## Your turn (solution)

```{r}
histogram <- function(df, var, ..., binwidth = NULL) {
  df |> 
    ggplot(aes(x = {{ var }})) + 
    geom_histogram(binwidth = binwidth, ...) + 
    labs(
      title = rlang::englue(
        "Histogram of {{ var }}, with binwidth {binwidth %||% 'default'}"
      )
    )
}

histogram(starwars, height, binwidth = 5)
```

## Mixing in other tidyverse functions

Your function can also include pre-processing of data.

```{r}
sorted_bars <- function(df, var) {
  df |> 
    mutate({{ var }} := {{ var }} |> fct_infreq() |> fct_rev()) |>
    ggplot(aes(y = {{ var }})) +
    geom_bar()
}

diamonds |> sorted_bars(clarity)
```

## Mixing in other tidyverse functions

Your function can also include pre-processing of data.

```r
sorted_bars <- function(df, var) {
  df |> 
    mutate({{ var }} := {{ var }} |> fct_infreq() |> fct_rev()) |>
    ggplot(aes(y = {{ var }})) +
    geom_bar()
}
```

:::{.incremental}
- If using `{{ }}` to specify a new column, use `:=`, not `=`.

- `fct_infreq()` reorders by decreasing frequency.

- `fct_rev()` reverses order, as y-axis starts at bottom.

- **Our turn**: let's try it
:::

## Summary (so far)

:::{.incremental}
- use embracing `{{}}` to interpolate bare column-names

  - function receiving the "embracing" has to be aware
  - look for `<data-masking>` in the help
  - use `rlang::englue()` to interpolate variables `{{}}` and values `{}`

- `...` is a useful "escape hatch" in function design:

  - put after required args, and before details
  - tell your users where the dots are going
:::

## Design and style

Use descriptive name, usually starts with a verb, unless it returns a well-known noun.

. . .

`mutate()`: verb, describes what it does

`median()`: noun, describes what it returns

## Order of arguments

- **required**: arguments without default values

- **dots**: can be passed on functions that your function calls

- **optional**: arguments with default values

. . .

Tidyverse Design:

 - [Position of dots](https://design.tidyverse.org/dots-after-required.html)
 
 - [Hadley's *first* Substack article](https://tidydesign.substack.com/p/argument-ordering)

## Order of arguments

Our histogram function:

```{r}
histogram <- function(df, var, ..., binwidth = NULL) {
  df |>
    ggplot(aes(x = {{var}})) +
    geom_histogram(binwidth = binwidth, ...)
}
```

- **required**: `df`, `var`

- **dots**: `...`

- **optional**: `binwidth`

## Order of arguments

```{r}
histogram <- function(df, var, ..., binwidth = NULL) {
  df |> 
    ggplot(aes(x = {{ var }})) + 
    geom_histogram(binwidth = binwidth, ...) 
}
```

Why *optional* after *dots*?

- user must name *optional* arguments, in this case `binwidth`.

- makes code easier to read when *optional* arguments used.

- more reasoning given in the [Tidyverse design guide](https://design.tidyverse.org/dots-position.html).

## Namespacing functions

Are we using `filter()` from `{dplyr}` or `{stats}`?

Three ways to sort this out:

:::{.incremental}
  - `library("conflicted")`, suitable for R scripts
  
  - `package::function()`, used in package functions

  - `#' @importFrom`, also used (sparingly) in packages
:::

## 📦 conflicted

```r
library("conflicted")
```

`{conflicted}` lets you know when you use a function that exists two-or-more packages that you've loaded.

. . .

To avoid conflicts, declare a preference: 

```r
# put in a conspicuous place, near the top of your script
conflicts_prefer(dplyr::filter)
```

## Your turn

In `functions-02-02-style.R`:

```r
library("tidyverse")

mtcars |> filter(cyl == 6)
```

- run it as-is
- add `library("conflicted")`, run again
- add a `conflict_prefer()` directive

## `package::function()`

This is the usual way when writing a function for a package:

```r
histogram <- function(df, var, ..., binwidth = NULL) {
  df |> 
    ggplot2::ggplot(ggplot2::aes(x = {{ var }})) + 
    ggplot2::geom_histogram(binwidth = binwidth, ...) 
}
```

:::{.incremental}
- makes it very clear where you are getting the function from
- can be verbose, especially if calling an external function often
:::

. . .

There is a balance to be struck.

## `#' @importFrom`

When you have a *lot* of calls to a given external function

Put this in `{packagename}-package.R`:

```r
#' @importFrom ggplot2 ggplot aes geom_histogram
NULL
```

. . .

Alternatively, from the R command prompt:

```r
usethis::use_import_from("ggplot2", c("ggplot", "aes", "geom_histogram"))
```

## `#' @importFrom`

```r
histogram <- function(df, var, ..., binwidth = NULL) {
  df |> 
    ggplot(aes(x = {{ var }})) + 
    geom_histogram(binwidth = binwidth, ...) 
}
```

Makes your code less verbose, but also less transparent

. . .

**To mitigate**: 

- put all your `@importFrom` in one conspicuous file: `{packagename}-package.R`
- use judiciously

## Design and style references

[R for Data Science](https://r4ds.hadley.nz/functions.html#style)

[Tidyverse Style Guide](https://style.tidyverse.org/functions.html)

[Tidyverse Design Guide](https://design.tidyverse.org/dots-position.html)

[Tidydesign Substack (Hadley)](https://tidydesign.substack.com/)

Also, look at tidyverse code at GitHub (my favorite is [{usethis}](https://github.com/r-lib/usethis/))

## Side effects

There are two kinds of functions:

. . .

**Pure function**:

- Returns a value that depends only on its inputs, e.g. `sum()`

. . .

**Uses side effects**:

- Depends on something other than inputs, e.g. `read.csv()`

- Or, makes a change in the environment, e.g. `print()`

## Pure function

```r
add <- function(x, y) {
  x + y
}
```

The return value depends *only* on the inputs.

Easier to test.

## Uses side effects

Side-effects can slow down your function:

 - it can be costly to read/write to disk, print to the screen.

. . .
 
Depending on side effects can introduce uncertainty:

 - are you *certain* of what `file.csv` contains?

. . .

Side effects aren't necessarily bad, but you need to take them into account:

  - need to take care when testing.

## Your turn

Discuss with your neighbor, are these function-calls are pure, or do they use side effects?

In `functions-02-03-side-effects.R`:

```r
x <- prod(1, 2, 3)

x <- print("Hello")

x <- ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point()

x <- sort(c("apple", "Banana", "candle"))
```

## Our turn: checking locale

Can be useful to consult `devtools::session_info()`: 

```{r}
# using `info = "platform"` to fit output on screen
devtools::session_info(info = "platform")
```

## Manage side effects using 📦 withr

Side effects can include:

:::{.incremental}
 - modifying environment: `Sys.setenv()`
 - modifying options: `options()`
 - setting random seed: `set.seed()`
 - setting working directory: `setwd()`
 - creating and wrting to a temporary file
:::

. . .
 
{withr} makes it a lot easier to "leave no footprints".

## Our turn: modifying locale

`sort()` uses locale (environment) for string-sorting rules 

```{r}
#| collapse: true
Sys.getlocale("LC_COLLATE")

sort(c("apple", "Banana", "candle"))
```

. . .

To temporarily set locale:

```{r}
#| collapse: true
withr::with_locale(
  new = list(LC_COLLATE = "C"),
  sort(c("apple", "Banana", "candle"))
)

Sys.getlocale("LC_COLLATE")
```

## Our turn: setting only within scope

```{r}
#| collapse: true
c_sort <- function(...) {
  # set only within function block
  withr::local_locale(list(LC_COLLATE = "C"))
  
  sort(...)
}

c_sort(c("apple", "Banana", "candle"))

Sys.getlocale("LC_COLLATE")
```

*Within curly brackets* applies to function blocks, it also applies to {testthat} blocks.

## Your turn

```{r}
library("testthat")

test_that("mtcars has expected columns", {
  expect_type(mtcars$cy, "double")
})
```

This passes, but R is doing partial matching on the `$`.

Modify the `test_that()` block to warn on partial matching.

You can get the current setting using:

```{r}
getOption("warnPartialMatchDollar")
```

Hint: use `withr::local_option()`.

## Your turn (solution)

```{r}
test_that("mtcars has expected columns", {
  
  withr::local_options(list(warnPartialMatchDollar = TRUE))
  
  expect_type(mtcars$cy, "double")
})
```

And yet...

```{r}
getOption("warnPartialMatchDollar")
```

## Summary

You can use tidy evaluation in {ggplot2} to specify aesthetics, add labels, and include {dplyr} preprocessing:

  - embracing `{{}}` for `<data-masking>` functions
  - be aware of `<tidy-select>` functions, work differently

. . .

<hr>

Using "standard" style and design makes things easier for you, your users, and *future you*.

. . .

<hr>

Be mindful of side effects, use {withr} to manage global state. 
