---
title: "Iteration: Part 2"
format: revealjs
---

## Reading multiple files

Using {purrr} to iterate can help you perform many tasks repeatably and reproducibly.

### Example

Read Excel files from a directory, then combine into a single data-frame.

## {here} package

When you first call `here::here()`, (simplified):

  - climbs your local directory until it finds a `.RProj` file
  - sets directory containing `.RProj` as reference-path
  - `here::here()` prepends reference-path to argument

If project in `/Users/ian/important-project/`:
 
```r
here("data/file.csv")
```
  
```
"/Users/ian/important-project/data/file.csv"
```  

## Our turn

In the `programming-r-exercises` repository:

- open `iteration-02-reading-files.R`
- restart R

## Our turn: reading data manually

Here's our starting code:

```r
data1952 <- readxl::read_excel(here("data/gapminder/1952.xlsx"))
data1957 <- readxl::read_excel(here("data/gapminder/1957.xlsx"))
data1962 <- readxl::read_excel(here("data/gapminder/1952.xlsx"))
data1967 <- readxl::read_excel(here("data/gapminder/1967.xlsx"))

data_manual <- bind_rows(data1952, data1957, data1962, data1967)
```

What problems do you see? 

(I see two real problems, and one philisophical problem)

Run this example code, discuss with your neighbor.

## Our turn: make list of paths

I see this as a two step problem:

- make a named list of paths, name is year
- use list of paths to read data frames, combine

Let's work together to improve this code:

```r
paths <-
  # get the filepaths from the directory
  fs::dir_ls(here("data/gapminder")) |>
  # extract the year as names
  # convert to list
  identity()
```

## Our turn: read data

```r
data <-
  paths |>
  # read each file from excel, into data frame
  # keep only non-null elements
  # set list-names as column `year`
  # bind into single data-frame
  # convert year to number
  identity()
```

## Handling failures

If we have a failure, we may not want to stop everything.

We can use a function-operator to return a modified function.

```{r}
#| error: true
library("readxl")
read_excel("not/a/file.txt")
```

<hr>

```{r}
#| error: true
library("purrr")
poss_rdxl <- possibly(read_excel, otherwise = NULL, quiet = TRUE)
poss_rdxl("not/a/file.txt")
```

<hr>

```{r}
#| message: true
poss_rdxl <- possibly(read_excel, otherwise = NULL, quiet = FALSE)
poss_rdxl("not/a/file.txt")
```

## Our turn: handle failure

In the `programming-r-exercises` repository:

- look at `data/gapminder_party/`
- try running your script using this directory

Create a new function:

```r
possibly_read_excel <- possibly(<you do the rest>)
```

Use this function in your script.

## If we have time

`list_rbind()` uses three different functional-programming techniques:

- *filters* out `NULL` values, `purrr::keep()`
- *maps* name of element to data-column, `purrr::imap()`
- *reduces* list to single data-frame, `purrr::reduce()`

```r
keep(\(x) !is.null(x)) |>
imap(\(d, name) mutate(d, year = name)) |>
reduce(rbind)
```

## Saving multiple outputs

## Extras

Three fundamental operations in functional programming

Given an array and a function:

 - `map()`:  make a new list, operating on each element
 - `filter()`: make a new list, subset of old list
 - `reduce()`: make a new "thing"

## dplyr using purrr?
 
We can use `map()`, `filter()`, `reduce()` to "implement", using purrr:

 - `dplyr::mutate()`
 - `dplyr::filter()`
 - `dplyr::summarise()`

I claim it's possible, I don't claim it's a good idea.

## Tabular data: two perspectives

- column-based: named list of column vectors

  ```json
  {
    mpg: [21.0, 22.8, ...],
    cyl: [6, 4, ...],
    ...
  }
  ```

- row-based: collection of rows, each a named list

  ```json
  [
    {mpg: 21.0, cyl: 6, ...}, 
    {mpg: 22.8, cyl: 4, ...}, 
    ...
  ]
  ```
  
## `dpurrr_filter()`

```{r}
library("tidyverse")

dpurrr_filter <- function(df, predicate) {
  df |>
    as.list() |>
    purrr::list_transpose(simplify = FALSE) |>
    purrr::keep(predicate) |>
    purrr::list_transpose() |>
    as.data.frame() 
}
```

<hr>

```{r}
dpurrr_filter(mtcars, \(d) d$gear == 3) |> head()
```

## `dpurrr_mutate()`

```{r}
dpurrr_mutate <- function(df, mapper) {
  df |>
    as.list() |>
    purrr::list_transpose(simplify = FALSE) |>
    purrr::map(\(d) c(d, mapper(d))) |>
    purrr::list_transpose() |>
    as.data.frame() 
}
```

<hr>

```{r}
mtcars |> 
  dpurrr_mutate(\(d) list(wt_kg = d$wt * 1000 / 2.2)) |> 
  head()
```

## `dpurrr_summarise()`

```{r}
dpurrr_summarise <- function(df, reducer, .init) {
  df |>
    as.list() |>
    purrr::list_transpose(simplify = FALSE) |>
    purrr::reduce(reducer, .init = .init) |>
    as.data.frame()
}
```

<hr>

```{r}
mtcars |> 
  dpurrr_summarise(
    reducer = \(acc, val) list(
      wt_min = min(acc$wt_min, val$wt), 
      wt_max = max(acc$wt_max, val$wt)
    ),
    .init = list(wt_min = Inf, wt_max = -Inf)
  )
```

## With grouping

First, a little prep work:

```{r}
ireduce <- function(x, reducer, .init) {
  purrr::reduce2(x, names(x), reducer, .init = .init)
}

summariser <- purrr::partial(
  dpurrr_summarise,
  reducer = \(acc, val) list(
    wt_min = min(acc$wt_min, val$wt), 
    wt_max = max(acc$wt_max, val$wt)
  ),
  .init = list(wt_min = Inf, wt_max = -Inf)
)
```

## Et voilà

```{r}
mtcars |> 
  split(mtcars$gear) |>
  purrr::map(summariser) |> 
  ireduce( 
    reducer = \(acc, x, y) rbind(acc, c(list(gear = y), x)),
    .init = data.frame()
  ) 
```

We can agree this presents no danger to dplyr.

In JavaScript, data frames are often arrays of objects (lists), so you'll see formulations like this (e.g. **tidyjs**).
